On Optimality of Finite Element Integration

### 2015-09-23 TOMS_Paper/assembly.tex
Rewrite bits of section 2 (the proof mostly)

\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into flop-optimal loop nests.
\item We comment on the cases in which such model does not lead to optimal loop nests. We will see that this might occur with particular forms that make extensive use of tensor algebra. We show, however, that our model still is capable of producing quasi-optimal loop nests.
\item We integrate the model with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.

@@

We prove that outer-product loop nests can be reduced to optimal form. The proof focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, which is beyond the scope of this paper, a more general treatise would be required to avoid sub-optimal outer loops in pathological cases).

\begin{Prop}
\label{prop:optimal-mln}
Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer-product loop nest with the body of $L_{i_{1}}$ being an independent basic block. Then an optimal synthesis $\Lambda'$ can be determined by eliminating sharing from both loops, starting from the loop with smallest sharing factor.
\end{Prop}
\begin{proof}
The demonstration is by construction and exploits linearity (recall that outer-product loop nests are multilinear by definition). We start with ``flattening'' the expressions appearing in the body of $L_{i_{1}}$; this is because, depending on the input structure, products might have to be recursively expanded in order to expose the space of all possible factorizations. The expanded expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).

We traverse the expression tree and determine the sharing factor of the two linear loops. Starting from the loop with smallest sharing factor, $L_i$, we incrementally factorize its symbols. Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other expression in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. This transformation is semantically correct, regardless of $L_i$ being $L_{i_{0}}$ or $L_{i_{1}}$: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about code motion). We repeat the same process for the loop with higher sharing factor, $L_j$.

We now have to demonstrate that the chosen factorization ordering, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true
\begin{enumerate}
\item Taking $L_j$, the loop with higher sharing factor, as point of departure leads to a sub-optimal $\Lambda'$ (optimal only in the best case). By contradiction, let us assume this is false. There are two possibilities: A) both $L_i$ and $L_j$ present sharing, but this occurs over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$. Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ is actually optimal (best case scenario). In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had taken $L_i$ as point of departure. Therefore, since $L^{sf}_{j} > L^{sf}_{i}$, the innermost loop results more expensive. We still wonder about the cost of hoisting, since clearly we are hoisting less expressions in this scenario. However, a hoisted symbol always costs one operation (a sum), while a non-factored outer product always costs two operations (a product and a sum). That is, factorization is always convenient. The initial assumption is therefore wrong.
\item Partial elimination of sharing leads to a sub-optimal $\Lambda'$ (optimal only in the best case). By contradiction, let us assume this is false. Because of linearity, we know that factorization eliminates sharing. We wonder what happens in the case of partial factorization; that is, at least one factorizable symbol, say along $L_i$, is left outside of the factored product (e.g., $a_i$ in $a_i(b_j + c_j + ...) + a_i e_j$). It is fairly simple to see that either there is a factorization opportunity along $L_j$, in which case no penalty needs be paid; or $\Lambda'$ is sub-optimal, since eventually the innermost loop will perform more operations. The initial assumption is therefore contradicted. Note that, if we were considering arities $n > 2$, this fact would generally be false.
\end{enumerate}
\end{proof}

For example, by applying this procedure to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.

@@

\subsection{Minimization of the Operation Count}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

>>>>

\begin{itemize}
\item We characterize loop nest optimality and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into optimal loop nests.
\item We comment on the cases in which such model does not lead to optimal loop nests. We will see that this might occur with particular forms that make extensive use of tensor algebra. We show, however, that our model still is capable of producing quasi-optimal loop nests.
\item We integrate the model with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

@@

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$. Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}. More in general, arbitrarily complex outer-product loop nests can systematically be reduced to optimal form. The constructive proof that we provide below focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).

\begin{Prop}
\label{prop:optimal-mln}
Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer-product loop nest with the body of $L_{i_{1}}$ being an independent basic block. Then an optimal synthesis $\Lambda'$ can be determined by eliminating sharing from both loops, starting from the loop with smallest sharing factor.
\end{Prop}
\begin{proof}
The demonstration is by construction and exploits linearity. We start with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_{i_{1}}$; this exposes the space of all possible factorizations. The resulting expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).

We traverse the expression tree and determine the sharing factor of the two linear loops. We take the symbols depending on the loop with smallest sharing factor, $L_i$, and we incrementally factorize them. Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$. The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about generalized code motion).

We now have to demonstrate that the chosen factorization strategy, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true (we do it by contradiction).
\begin{enumerate}
\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$. Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario). In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$. However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product. Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). Let us assume, for a moment, that this is false. In this scenario, at least one symbol is excluded from the factored product. Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$. It is fairly obvious to see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal. Consequently, the hypothesis is generally contradicted.
\end{enumerate}

Note that if we were considering outer-product loop nests of arity $n > 2$, unless relaxing Definition~\ref{def:mln-optimality}, we would need a much more complex construction to guarantee optimality. 
\end{proof}

@@

\subsection{Operation Count Minimization}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.
